\documentclass[article]{jss}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{url}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{color}
\usepackage{natbib}

%% Fallbacks so the paper also builds against a minimal jss.cls (the real JSS
%% class defines all of these). Harmless when the full class is present.
\providecommand{\proglang}[1]{\textsf{#1}}
\providecommand{\pkg}[1]{\textbf{#1}}
\providecommand{\code}[1]{\texttt{#1}}

% Let TeX relax interword spacing slightly to avoid minor margin protrusions
% on lines densely packed with \pkg/\code tokens.
\setlength{\emergencystretch}{3em}

%% JSS article information
\author{
  Peter Cotton\\
  Microprediction\\
  \email{peter.cotton@microprediction.com}\\
  \url{https://skaters.microprediction.org}
}

\title{Transforms All the Way Down --- Automatic Online Distributional Forecasting by Conjugation}

\Plaintitle{Transforms All the Way Down - Automatic Online Distributional Forecasting by Conjugation}
\Shorttitle{Transforms All the Way Down}

\newcommand{\skatersabstract}{
The \proglang{Python} package \pkg{skaters} is an online, distributional,
univariate time-series forecaster built by \emph{conjugation}: an invertible
transform wraps an inner
forecaster --- pushing each observation forward, forecasting in the transformed
space, and pulling the whole predictive distribution back through the inverse.
Because the inner forecaster is itself a conjugation, models nest
\emph{transforms all the way down} onto a single distributional leaf, and an
ensemble is just another forecaster that combines forecasters. The leaf fits its
shape by optimising a proper
scoring rule, so the objective is a choice rather than a fixed property of the
method: pointed at likelihood the recursion models the conditional density,
pointed at CRPS it is competitive with the CRPS specialists on their own metric
--- fitting and tail-shaping are two settings of one machine, not two machines. The library is written twice, in pure \proglang{Python} and in
zero-dependency \proglang{JavaScript} that agrees to within $10^{-6}$, so a model
runs unchanged on a server or in a browser. \pkg{skaters} was first envisaged as a \emph{collection} of forecasters
differing by prior; the study reported here pushed the other way, collapsing
that collection into a single forecast function with no exposed tuning
parameters, which we call \code{laplace} after the weak (indifference) prior it
places over its candidate pool. Positioned as a general-purpose forecaster for
economic and other non-price time series, it leads the per-series held-out
log-likelihood race against classical, neural, and pretrained foundation-model
baselines on \proglang{FRED} series; on asset-price and return series a
\pkg{GARCH}-$t$ model remains the better choice, and we report that split
rather than average it away.
}

\begin{document}

\maketitle

\begin{abstract}
\skatersabstract
\end{abstract}

\noindent\textbf{Keywords:} probabilistic forecasting, time series, proper
scoring rules, conformal prediction, online learning, \proglang{Python},
\proglang{JavaScript}, \pkg{skaters}.

%% -------------------------------------------------------------------------
\section{Introduction} \label{sec:intro}

Most deployed forecasters return a point and, perhaps, an interval bolted on
afterwards. Yet the quantity a decision-maker actually needs is the
\emph{predictive distribution}: the mean for an estimate, the spread for risk,
the tails for position sizing, and a density for likelihood-based evaluation and
combination. Two strands of recent work take distributions seriously but each
gives something up.

Classical state-space and exponential-smoothing models
\citep{hyndman2002state,hyndman2008automatic} are distributional but rarely
\emph{online and composable}: adding a transform usually means re-deriving the
filter. Conditional-heteroscedasticity models \citep{engle1982autoregressive,
bollerslev1986generalized} capture the heavy tails and volatility clustering of
financial data, but commit to a single parametric form and are refit in batch.
At the other extreme, conformal prediction
\citep{vovk2005algorithmic,vovk2019conformal} is distribution-free and online,
but a conformal predictive system outputs a cumulative distribution
function, not a density. Its finite-sample guarantee is \emph{marginal
coverage} --- not CRPS optimality --- and its native output (a set, interval,
quantile family, or CDF) carries no unambiguous Lebesgue density, so
log-likelihood scoring requires additional smoothing choices on which the
resulting likelihood then depends; an unsmoothed empirical conformal CDF
assigns zero density --- and hence $-\infty$ log-likelihood --- to any value
outside the observed residual range. Modern neural and foundation models
\citep{salinas2020deepar,ansari2024chronos,das2024timesfm,woo2024moirai} are
powerful distributional learners, but are built for global cross-series training
on hardware accelerators, not for a single univariate stream running in a
browser.

\pkg{skaters} takes a third route. Every node --- transform, ensemble, or leaf
--- is a function \code{(y, state) -> ([Dist], state)} that consumes one
observation and emits a list of \code{Dist} objects, one per forecast horizon,
where a \code{Dist} is a weighted Gaussian mixture. Because the type is closed
under composition, an entire model is a stack of invertible transforms ---
transforms all the way down --- with one distributional leaf at the bottom, and
ensembles are simply skaters that combine other skaters. This single abstraction
yields three contributions developed in turn:

\begin{enumerate}
\item \emph{A pluggable objective} (Sections~\ref{sec:trunk}--\ref{sec:leaf}).
  The leaf fits its mixture weights by a proper scoring rule. Pointed at
  log-likelihood it models the conditional distribution directly; with the
  terminal leaf instead pointed at CRPS while the likelihood-weighted trunk is
  unchanged, the same forecaster becomes competitive with a CRPS specialist on
  CRPS and stronger on likelihood --- the objective is a knob at the base of the
  recursion, not a property of the method.
\item \emph{Distributional structure that survives composition}
  (Sections~\ref{sec:lattice}--\ref{sec:doob}): a mean-preserving lattice
  projection that places atoms on the exact values a series revisits without
  moving the ensemble mean and vanishes on continuous data; an online coordinate
  search over whether a series is simple in a log, root, or linear coordinate;
  composable volatility and mean-reversion transforms (a \pkg{GARCH}-$t$ leaf, an
  Ornstein--Uhlenbeck group) that specialise the one forecaster without adding a
  second; and a multi-scale granularity mixture (Section~\ref{sec:granularity})
  that extends the same likelihood weighting to the sampling clock itself for
  multi-step horizons.
\item \emph{A benchmark against twelve distributional baselines}
  (Section~\ref{sec:experiments}), chosen to be difficult to beat on each
  metric:
  the heavy-tail state of the art (\pkg{GARCH}-$t$, both \pkg{arch} and
  \pkg{rugarch}), the reference \proglang{R} forecasters (the real Hyndman
  \pkg{auto.arima}, \pkg{thetaf}, Svetunkov's \pkg{ADAM}, \pkg{nnetar}), and
  \pkg{AutoARIMA} paired with split- and adaptive-conformal residuals plus
  conformal seasonal pools --- the CRPS-optimised methods hardest to beat on that
  metric. Every method is scored on held-out log-likelihood and CRPS through a
  single, identical \code{Dist}-based protocol.
\end{enumerate}

The unifying premise is the No-Free-Lunch theorem \citep{wolpert1997no} taken
seriously: every ``method'' is a prior, and averaged over all series none
dominates. The right response is not to pick one but to make the priors
explicit and weight them online: each candidate acts as an \emph{operational
prior} over series behaviour, and likelihood weighting limits the practical cost
of a poor one by draining its weight as evidence accumulates.

No-Free-Lunch also fixes the scope of our claim. We position \pkg{skaters} as a
general-purpose forecaster for economic and other non-price series, and that is
where the benchmark of Section~\ref{sec:experiments} is run. Asset-price and
return series are the hard case --- but not because of heavy tails, which the
scale-mixture leaf (Section~\ref{sec:leaf}) already models. They are hard because
they are close to \emph{martingales}: the conditional mean carries almost no
signal, so what forecasting skill remains lives in the conditional \emph{variance},
where the dedicated recursion of a \pkg{GARCH}$(1,1)$-$t$ still tracks volatility
clustering better than our online scale. There we do not claim to win, recommend
\pkg{GARCH}-$t$, and report the result as a split rather than dilute it into a
pooled average. The distinction is not the price label but the martingality axis
itself (Section~\ref{sec:martingality}): prices are merely its random-walk-heavy
extreme, and the \pkg{GARCH}-$t$ contest is governed by where a series falls along
it. The library's
conditional-variance building blocks (Section~\ref{sec:doob}) narrow but do not
close that gap (Table~\ref{tab:price}: the \pkg{GARCH}-$t$ leaf halves the
paired log-score deficit on $2{,}500$ price series); matching \pkg{GARCH}-$t$ on price series with an online,
dependency-free recursion is future work, not a present claim.

%% -------------------------------------------------------------------------
\section{Using the package} \label{sec:software}

\pkg{skaters} is pure \proglang{Python} with no required dependencies
(\code{pip install skaters}); the \proglang{JavaScript} twin lives in the same
repository and loads as an ES module. The entire public surface is one factory
and one value type:

\begin{verbatim}
from skaters import laplace

f = laplace(k=20)          # the general forecaster; multi-scale at k > 1
state = None
for y in stream:
    dists, state = f(y, state)     # list[Dist], one per horizon 1..k
    d = dists[0]
    d.mean, d.std              # point forecast and spread
    d.quantile(0.975)          # tail quantile
    d.logpdf(y), d.crps(y)     # proper-score a realised value
\end{verbatim}

\noindent Every model in the paper is this call with at most one knob changed:
\code{laplace(k, objective="likelihood")} repoints the terminal leaf
(Section~\ref{sec:leaf}), \code{laplace(k, sticky=False)} disables the lattice
projection (Section~\ref{sec:lattice}), \code{laplace(k, leaf=garch\_leaf)}
swaps in the conditional-variance leaf (Section~\ref{sec:doob}), and
\code{laplace(k, scales=[1])} opts out of the multi-scale mixture
(Section~\ref{sec:granularity}). The building blocks (\code{conjugate}, the
transforms, the leaves, the ensembles) are exported for composition but no
benchmark below uses them directly.

\paragraph{The two recursions, exactly.}
Writing $T$ for a transform with state $s_t$ (Section~\ref{sec:dist}) and $g$
for the inner skater, \emph{conjugation} is
\begin{align*}
(z_t,\; s_{t+1}) &= T.\code{forward}(y_t,\; s_t) \\
(D'_{1:k},\; u_{t+1}) &= g(z_t,\; u_t) \\
D_{1:k} &= T.\code{inverse\_k}(D'_{1:k},\; s_{t+1}),
\end{align*}
and the \emph{terminal-leaf ensemble} over candidates $g_1,\dots,g_n$ with
depths $d_i$ updates, per observation,
\begin{align*}
\ell_i &\leftarrow \gamma\,\ell_i + \eta \log p_i(y_t) - c\,d_i,
  \qquad w_i \propto e^{\ell_i}, \\
\hat\mu_h &= \textstyle\sum_i w_i\, \mu_{i,h}, \qquad
r_t = y_t - \hat\mu_{1}^{(t-1)}, \qquad
D_h = \code{leaf}_h(r_t) \text{ shifted by } \hat\mu_h,
\end{align*}
with $\gamma = 0.99$ (forgetting), $\eta = 0.8$ (learning rate), $c = 0.005$
(per-depth complexity penalty), and $p_i$ the candidate's resolved one-step
predictive. These constants are fixed defaults, not exposed parameters; they are
the same for every series in every experiment below.

\paragraph{Reproducibility.}
Section~\ref{sec:heritage} lists the exact commands; every table is regenerated
by one harness invocation from the committed result files, and the
\proglang{Python}/\proglang{JavaScript} parity suite (roughly $100{,}000$ probe
values at $10^{-6}$) runs in CI.

%% -------------------------------------------------------------------------
\section{The \code{Dist} abstraction and composable transforms} \label{sec:dist}

A \code{Dist} is a weighted mixture $\sum_i w_i\,\mathcal N(\mu_i,\sigma_i^2)$
with $\sigma_i>0$ and $\sum_i w_i = 1$. It exposes \code{mean}, \code{std},
\code{pdf}, \code{cdf}, \code{logpdf}, \code{quantile}, and a closed-form
\code{crps}, together with the affine maps (\code{shift}, \code{scale},
\code{affine}) and a moment-matching \code{prune} needed to propagate it through
transform inverses and to bound component growth in ensembles. The Gaussian
mixture is a deliberate choice: it is closed under affine maps and mixing, admits
a closed-form CRPS through the elementary identity
$\mathbb{E}\,|\mathcal N(m,s^2)| = 2 s\,\phi(m/s) + m\,(2\Phi(m/s)-1)$, and
approximates heavy tails arbitrarily well as a scale mixture.

A \emph{transform} is a stateful causal map
\[
(z_t,\; s_{t+1}) \;=\; T(y_t,\; s_t),
\]
whose \code{forward} consumes one observation and the stored state and emits one
transformed value, paired with an \code{inverse\_k} that reconstructs predictive
distributions for future $y$'s from $k$ \code{Dist}s in the transformed space,
\emph{conditional on the current transform state} $s_t$. Several of these maps
(differencing, standardisation, autoregression) are not scalar bijections on
$\mathbb R$; they are invertible as maps on paths given the stored state, which
is exactly what \code{inverse\_k} receives. Differencing, exponential smoothing,
drift, Holt's linear trend, fractional differencing, autoregression (via online
recursive least squares), \pkg{GARCH}-style scaling, seasonal differencing,
power and Yeo--Johnson coordinates \citep{box1964analysis,yeo2000new}, and
standardisation are all transforms. \emph{Conjugation} composes a transform with
an inner skater,
\[
y \;\xrightarrow{\;T\;}\; y' \;\to\; \text{inner} \;\to\; D'
  \;\xrightarrow{\;T^{-1}\;}\; D ,
\]
and because every step is online and the inverse propagates the full mixture,
multi-step uncertainty accumulates through the inverse. One caveat is stated
once and holds throughout: the emitted list contains one \emph{marginal}
predictive per horizon, not a joint law over the path. Linear inverses propagate
means exactly and propagate variances under the library's recursive
independent-innovation convention --- the differenced inverse, for example, adds
the marginal innovation variances, $\sum_{h}\sigma_h^2$, which is exact when
future transformed innovations are conditionally uncorrelated and an
approximation otherwise. The convention is audited rather than merely stated:
the returned state carries per-horizon calibration diagnostics --- the
probability integral transform of each arriving observation under the
prediction issued $m$ steps earlier \emph{for it}, and its standard-normal
quantile $z$ --- and the empirical standard deviation of the horizon-$m$ $z$
sequence estimates precisely the factor by which the independent-innovation
convention mis-states the $m$-step scale. (Even under perfect calibration the
horizon-$m$ $z$ values are $(m{-}1)$-dependent, from overlapping forecast
windows, so consecutive values are not independent evidence.) The inner skater is unconstrained: it may
itself be a conjugation, or an ensemble of conjugations, so a model is
\emph{transforms all the way down}, bottoming out at a single distributional leaf
--- the one non-transform in the stack.

Structurally this is the recipe of a normalizing flow
\citep{rezende2015variational} --- a base density pushed through invertible maps
--- but the regime is inverted. A flow \emph{learns} its coupling layers by
stochastic gradient descent over a batch; here the transforms are \emph{fixed and
analytically invertible} as scalar maps (for nonlinear coordinates the
\emph{pushforward of the predictive} uses a delta-method linearisation,
Section~\ref{sec:coord}), their few parameters adapting \emph{online} by
score-driven, Tweedie-style updates \citep{creal2013generalized} one observation
at a time, in $O(1)$, with no training phase and no gradients. The base is a
weighted mixture with a pluggable proper-scoring objective, not a fixed Gaussian.
It is a flow's compositional structure carried into a streaming, closed-form,
univariate-forecasting regime.

\paragraph{Numerical note.}
Mixture variance is computed in centred form,
$\operatorname{Var} = \sum_i w_i\,(\sigma_i^2 + (\mu_i-\bar\mu)^2)$, rather than
$\mathbb E[X^2]-\mathbb E[X]^2$. The latter catastrophically cancels when a tight
component sits at a large mean, silently producing $\sigma=0$ and an invalid
predictive. Running moments use Welford's recurrence \citep{welford1962note}.

%% -------------------------------------------------------------------------
\section{The terminal-leaf ensemble: modelling the mean} \label{sec:trunk}

Bayesian model averaging over a heterogeneous pool
\citep{hoeting1999bayesian,raftery2005using} combines full predictive
distributions. In this implementation that dilutes the very shape the leaf is
for: with dozens of candidates the pooled mixture is dominated by the dispersion
of the candidate \emph{means} (and by component pruning), so a carefully shaped
heavy-tailed leaf inside the pool has little influence on the output tail. We therefore use the candidate models
only as mean forecasters, weight them online by predictive likelihood
(with a learning-rate shrinkage and a per-depth complexity penalty, in the
spirit of gradient boosting),\footnote{The penalty is charged per step ---
rent, not a prior. Under geometric forgetting of the accumulated log-evidence
(the shipped policy discounts at $\lambda=0.99$) the discounted rent
telescopes to a bounded handicap proportional to $\text{depth}/(1-\lambda)$,
i.e.\ an MDL-style prior on model complexity; at $\lambda=1$ it grows linearly
and would eventually evict any deep candidate regardless of skill. The
forgetting factor is what makes the complexity penalty a prior rather than a
bias.} combine their means, and model the distribution of
the combined residual with a single terminal leaf:
\[
 y \;\to\; \hat\mu = \sum_i w_i\,\hat\mu_i, \qquad
 r = y-\hat\mu \;\to\; \text{leaf} \;\to\; D, \qquad
 \hat D = D \text{ shifted by } \hat\mu .
\]
Because exactly one leaf reaches the output, its shape --- heavy tails, learned
online --- survives undiluted. The trunk is a likelihood-weighted ensemble whose
only job is to get the conditional mean right; the shaping of the distribution is
left to the base of the recursion.
The candidate pool spans exponential smoothing, differencing, drift, Holt's
linear trend, autoregression, grouped long-lag autoregression, fractional
differencing, a seasonal block, a Yeo--Johnson coordinate grid
(Section~\ref{sec:coord}), and a fast/slow two-systems block.

%% -------------------------------------------------------------------------
\section{Metric-agnostic leaves: a pluggable base} \label{sec:leaf}

The leaf is itself a fixed Gaussian \emph{scale mixture} --- components
$\mathcal N\!\big(0, (c_k\,\hat\sigma)^2\big)$ over a fixed log-spaced
dictionary $\{c_k\}$ of standard-deviation multipliers, with $\hat\sigma$ an
exponentially weighted scale --- whose only learned quantities are the mixture
weights $\{w_k\}$ on the simplex. Student-$t$ laws are particular Gaussian
scale mixtures (an inverse-gamma mixing law over the variance), and the finite
log-spaced dictionary approximates such heavy-tailed scale mixtures while
remaining a plain \code{Dist}. The weights are fit online by
a proper scoring rule \citep{gneiting2007strictly}:
\begin{itemize}
\item \code{likelihood} --- recency-weighted expectation--maximisation on the
  component responsibilities;
\item \code{crps} --- exponentiated-gradient descent on the simplex minimising the
  closed-form mixture CRPS \citep{matheson1976scoring}; the gradient is exact,
  since the mixture CRPS is a finite sum of $\mathbb E\,|\mathcal N(m,s^2)|$ terms.
\end{itemize}

Because the leaf is where the whole conjugation bottoms out, it is exactly where a
downstream metric should get its vote. The trunk's job is to model, and the right
objective for modelling is likelihood, a proper, tail-sensitive score that is,
up to a constant, the Kelly log-growth criterion \citep{kelly1956new}; corrupting
it to chase CRPS would surrender the density, as conformal systems do by
construction.
Pointing only the terminal leaf at CRPS shapes the residual tail without touching
the model above it. Section~\ref{sec:ablations} shows this configuration
improving on a pure-likelihood policy on CRPS and on likelihood alike: a CRPS-fit
leaf is more outlier-robust, so it even generalises slightly better in
log-likelihood.

There is also a well-posedness asymmetry between the two objectives.
Minimising expected log-score projects the truth onto the leaf family in
Kullback--Leibler divergence, which for a location--scale family amounts to
moment matching and requires a finite second moment. Minimising expected CRPS
projects in the Cram\'er distance $\int (F-G)^2$, and $\mathrm{CRPS}(F,y)$ is
finite whenever $F$ has a first moment \citep{gneiting2007strictly}. On
fat-tailed streams the CRPS leaf therefore remains well-defined in regimes
where variance matching does not --- a theoretical companion to the robustness
observed empirically.

%% -------------------------------------------------------------------------
\section{The lattice projection} \label{sec:lattice}

Administrative and grid-quoted series --- policy rates, posted prices, pegged
exchange rates --- revisit a small set of exact values, and a continuous
predictive density cannot place mass there. We add a \emph{projection}: a
recency-weighted frequency table of exact values, with near-Dirac atoms on the
values revisited above a noise floor, each carrying its observed frequency as
probability. The atom is mean-preserving: the continuous part is
recentred so the atoms add mass without moving the ensemble mean.

Two properties make this a safe default. First, on continuous data nothing is
revisited, no atom fires, and the projection vanishes identically --- it is free
when unused. Second, unlike a last-value spike it captures values that recur
often but never consecutively (a rate that repeatedly returns to a round number).
The pull of the level (its tendency to persist) is left to the trunk, where the
random-walk candidate earns its weight by likelihood, so the projection remains a
pure statement about mass, not mean.

\paragraph{Scoring convention.}
Because the \code{Dist} stays a Gaussian mixture, an ``atom'' is a narrow
Gaussian, not a true Dirac: its standard deviation is scale-adaptive,
$\max(0.005\,\sigma_D, 10^{-9})$ with $\sigma_D$ the predictive's own std, its
mass is the value's observed recency-weighted revisit frequency (total atom mass
capped at $0.999$), and a value qualifies only once revisited above a noise
floor ($1.8\times$ the table's EMA rate), never on first sight. The density is
therefore an ordinary Lebesgue density and every method is scored by the same
rule. The convention matters for log-likelihood: at a revisited exact value the
score grows like $\log(1/\text{width})$, so the projection is a bandwidth-%
dependent log-score machine on grid series. That is precisely why
Table~\ref{tab:sota} restricts to \emph{continuous} series (under $5\%$
repeating changes), where the projection fires rarely and its contribution to
the reported win-rates is immaterial. On grid-heavy series we measured the
sensitivity directly: sweeping the atom width over two orders of magnitude
($0.0005$--$0.05$ of the predictive std, $400$ grid-heavy series) moves the
median held-out log-likelihood from $+7.8$ to $+5.1$ (off: $+3.2$) --- the
magnitudes are bandwidth artefacts, as the scaling predicts --- while the
per-series \emph{winner} against \pkg{AutoETS} agrees on $91\%$ of series
between the extreme widths (win-rate $66\%$ vs $57\%$), and CRPS, being
bounded, barely moves ($0.0035$ vs $0.0053$). Grid-series log-likelihood
magnitudes should be read as qualitative; the orderings survive the bandwidth
choice. The convention has one further consequence, this time favourable:
evaluating the mixture cdf at a revisited value $v$ carrying atom mass $p$
yields, as the atom width shrinks, $F(v^-)+p/2$ --- the nonrandomised
\emph{mid-PIT} recommended for discrete data \citep{czado2009predictive} ---
so probability-integral-transform diagnostics remain centred on lattice series
without any explicit discrete-data correction.

%% -------------------------------------------------------------------------
\section{Coordinate learning} \label{sec:coord}

Whether a series is simple in a linear, log/multiplicative, or root coordinate is
itself learnable. We add a coarse grid of Yeo--Johnson $\lambda$
candidates --- the signed Box--Cox family \citep{box1964analysis,yeo2000new},
defined on all of $\mathbb R$ --- to the pool, so the ensemble learns the
coordinate online rather than committing up front. This is No-Free-Lunch-safe: a
wrong coordinate is simply down-weighted by likelihood. The same mechanism
expresses a non-negativity prior at $\lambda=0$ (the log coordinate), though we
note that because the \code{Dist} is a Gaussian mixture the inverse uses the
delta-method linearisation, so non-negativity is approximate rather than exact.

%% -------------------------------------------------------------------------
\section{Volatility and mean reversion as composable transforms} \label{sec:doob}

The same compositional vocabulary expresses conditional volatility and mean
reversion without a separate forecaster. A \pkg{GARCH}-style scaling transform
divides by a learned conditional standard deviation; an Ornstein--Uhlenbeck
transform (\code{ou\_transform}) reverts the level toward a running mean at a
learned speed; and a \pkg{GARCH}$(1,1)$-$t$ terminal leaf (\code{garch\_leaf})
gives the conform-last stage a conditional-variance recursion with Student-$t$
tails. These are building blocks: \code{laplace} already carries an OU group in
its multi-step ($k>1$) candidate pool, and \code{laplace(leaf=garch\_leaf)} swaps
in the conditional-variance leaf for heavy-tailed series --- no second named model
required. (On asset-price/return series the right tool remains a fitted
\pkg{GARCH}-$t$; see the price caveat of Section~\ref{sec:experiments}.)

These volatility recursions have a Bayesian reading. Each is a \emph{score-driven}
update \citep{creal2013generalized,harvey2013dynamic}: the \pkg{GARCH} variance
recursion $h_t = h_{t-1} + (1-\delta)(y_t^2 - h_{t-1})$ is, term for term, the
inverse-Fisher-scaled conditional score, and \citet{hansen2026tweedie} show via
Tweedie's formula that under a conjugate prior and local precision discounting it
is the \emph{exact} Bayesian posterior-mean correction for the variance. The same
analysis identifies the smoothing factor $\alpha = 1-\delta$ that governs the
exponential-smoothing and \pkg{GARCH}-style transforms of
Section~\ref{sec:dist}, and recovers the Gaussian-location update as the Kalman
filter. Averaging the candidates therefore blends a family of (approximate)
Bayesian volatility filters, a complementary justification for treating the clock
itself as the object to learn.

%% -------------------------------------------------------------------------
\section{Granularity: mixing time scales} \label{sec:granularity}

Multi-step forecasting exposes one more axis a forecaster silently commits to:
the \emph{sampling granularity}. Natively, \code{laplace(k)} reaches horizon $h$
by fanning its one-step model out $h$ steps, and on long horizons that fan-out
can inflate the predictive variance badly. A model running on a \emph{decimated}
clock --- every $s$-th observation --- reaches the same fine horizon $h$ in only
$\mathrm{round}(h/s)$ of its own steps, so its predictive stays tight; the price
is blindness to the skipped points. Neither clock dominates: empirically
(Section~\ref{sec:multistep}) the fine scale wins short horizons and the coarse
scale long ones. The response is the library's usual one --- do not choose. The
\code{multiscale} wrapper --- the default inside \code{laplace} whenever $k>1$
(opt out with \code{scales=[1]}) --- runs a full
\code{laplace} on each of several decimated clocks (default strides $\{1,
\lceil\sqrt{k}\rceil, k\}$) and, per horizon, mixes their predictive
distributions with \code{Dist.combine} under likelihood softmax weights, so each
horizon selects its effective granularity through the weights exactly as the
pool selects coordinates or differencing orders.

Granularity is, however, the one candidate axis that cannot live \emph{inside}
the pool, which is why it is a wrapper rather than a transform. Decimation is
many-ticks-to-one, breaking the scalar-in/scalar-out contract of
\code{conjugate}; and a coarse-clock model emits no fine one-step predictive,
so the trunk's one-step likelihood weighting (Section~\ref{sec:trunk}) has
nothing to score it by. Two details matter. \emph{Phase}: a fixed-phase stride-$s$
model is up to $s-1$ ticks stale at forecast time, so the wrapper keeps $s$
phase-shifted copies per scale, of which exactly one steps per tick --- the
freshest coarse forecast is always anchored at the current tick, at a per-tick
cost of one base step per scale regardless of $s$. \emph{Degeneracy}: horizon
$h$ draws only on scales $s \le h$, so at $k=1$ the wrapper is exactly
\code{laplace(1)} --- multi-scale is free where it has nothing to add.

%% -------------------------------------------------------------------------
\section{Honest tails: the conditional tail fit} \label{sec:tails}

A density forecaster makes quantitative promises, and the deepest ones are
tail probabilities. Measured prequentially on the study universe, the
library's own promises initially failed there: the calibration $z$ of
Section~\ref{sec:software} was honest in the bulk (its nominal central 90\%
interval covered 89.2\%) but its Gaussian read fired at roughly eight times
the stated rate at tail probability $10^{-3}$ --- the predictive tails were
too thin beyond three sigma. The repair cannot be a tail-weighted refit: the
weighted log score $w(y)\log f(y)$ is improper, rewarding densities that
inflate the weighted region \citep{gneiting2011comparing}. The consistent
objective is the \emph{censored likelihood} \citep{diks2011likelihood} ---
full $\log f(y)$ inside a region, $\log \Pr(\text{outside})$ otherwise ---
whose extra term is forced by a theorem: no proper score assesses tail shape
without also charging for tail mass \citep{brehmer2019tails}.

Since v0.13.0 every \code{laplace} predictive therefore carries
generalized-Pareto tails --- peaks-over-threshold in the spirit of
\citet{siffer2017anomaly}, applied to the forecaster's own surprise stream.
The body defines a \emph{predictable} tail region (its calibration $z$
beyond thresholds frozen at warm-up quantiles, fixed before each
observation); a trailing tail model fits the exceedances per side by
censored maximum likelihood, which factorizes into an exceedance rate (an
EWMA, forgetting like every other estimator in the library) and a
conditional GPD shape (Grimshaw profile). Each issued \code{Dist} is spliced
--- body density in the interior, rescaled so mass is exact; GPD beyond ---
and the splice sits \emph{below} the calibration state, so \code{logpdf},
\code{cdf}, \code{quantile}, CRPS and \code{state["z"]} all read the
corrected tail. Information flows body $\to$ region $\to$ tail $\to$ output
and never backwards; \code{tails="gaussian"} opts out. Intake is winsorised
at the fitted $10^{-3}$ conditional excess with a changepoint escape, so an
anomaly cluster cannot fatten the fit and mask its successors.

Measured prequentially against \code{tails="gaussian"} on the same series:
held-out log-likelihood improves by ${\sim}+0.03$ nats per tick on 96--98\%
of series --- non-price and price alike, at one-step and multi-step horizons
--- while CRPS, the default leaf objective, is unchanged to within noise.
Median empirical coverage of nominal central intervals (142 non-price
series):

\begin{center}
\begin{tabular}{lcc}
\hline
nominal & Gaussian tails & GPD tails (default) \\
\hline
90\%   & 89.17\% & 90.08\% \\
99\%   & 97.62\% & 98.93\% \\
99.9\% & 99.13\% & 99.85\% \\
\hline
\end{tabular}
\end{center}

The Gaussian 99.9\% interval breached its budget $8.7\times$; the spliced
one is within 0.05 points of nominal, and a stated alarm budget now
converts directly: alarm when $\mathrm{erfc}(|z|/\sqrt{2}) < \alpha$, with
measured rate ${\approx}1.4\times10^{-3}$ at nominal $10^{-3}$, the residue
consistent with the genuine anomalies real series contain. The findings,
the propriety construction, and the negative results that scope the claim
--- tail \emph{quantile} location moves little, and near-deterministic
waveform streams remain out of scope --- are written up in the repository
(\code{papers/when-to-alarm.md}).

%% -------------------------------------------------------------------------
\section{Experiments} \label{sec:experiments}

\paragraph{Universe.}
To avoid a hand-picked series list we draw from the Federal Reserve Economic Data
\citep{fred}, taking cached daily series and auto-transforming each to its
one-step \emph{changes} (log-difference for strictly positive levels, else first
difference) --- the stationary-ish innovation stream where heavy tails and regime
shifts live. The rule is part of the shared task definition: every method, ours
and the baselines', forecasts the \emph{same} change stream. Two caveats. Scale
comparability of absolute log-likelihoods holds only on the log-differenced
(unitless) subset --- first differences keep the series' units, and a rescaling
$y \mapsto ay$ shifts the log-density by $-\log a$ --- so cross-series averages
of absolute scores are indicative at best, and all win-rates and paired score
differences are computed \emph{within} series, where the Jacobian cancels.
And because the log-vs-linear choice is a coordinate decision made before any
model sees data, we checked its effect directly: rerunning the contest on $250$
strictly-positive continuous series with plain first differences for every
method moves \code{laplace}'s log-likelihood win-rates from $94/94/70\%$ to
$98/96/72\%$ (vs \pkg{AutoARIMA}/\pkg{AutoETS}/\pkg{GARCH}-$t$) --- the rule
does not manufacture the edge; if anything it understates it. We keep the last $1000$ changes per series and score the last $300$ as a
rolling one-step-ahead test window after a burn-in. Consistent with the scope set
in Section~\ref{sec:intro}, the evaluation universe is the \emph{non-price}
economic series: we classify each series a priori from its FRED title and exclude
the asset-price classes (equity, foreign exchange, commodity), whose return
series belong to the conditional-volatility regime and are treated separately in
the price caveat below. The non-price universe is the general economic forecasting
problem \pkg{skaters} is built for.

\subsection{Against twelve distributional baselines} \label{sec:sota}

On the non-price universe we compare the general forecaster \code{laplace}
against the distributional one-step baselines we could run on CPU at this
scale. From
\pkg{statsforecast} \citep{garza2022statsforecast}: \pkg{AutoARIMA} and
\pkg{AutoETS}. From the \proglang{R} forecasting ecosystem --- the reference
implementations, not Python ports --- the actual Hyndman \pkg{auto.arima} and
\pkg{thetaf} (the \pkg{forecast} package \citep{hyndman2008automatic}; the Theta
method \citep{assimakopoulos2000theta} won the M3 competition), Svetunkov's
\pkg{ADAM} auto state-space model \citep{svetunkov2023adam}, a neural
autoregression (\pkg{nnetar}), and a \pkg{rugarch} \pkg{GARCH}$(1,1)$-$t$
\citep{ghalanos2022rugarch}. From \pkg{arch}, a constant-mean \pkg{GARCH}$(1,1)$
with Student-$t$ innovations
\citep{bollerslev1986generalized,bollerslev1987conditionally}, the classical
heavy-tail / volatility-clustering state of the art. And four conformal opponents:
\pkg{AutoARIMA} paired with split-conformal residual quantiles
\citep{vovk2019conformal} and its adaptive variant (ACI,
\citealp{gibbs2021adaptive}), plus conformal seasonal pools (CSP,
\citealp{manokhin2026csp}) in a fixed and a coverage-adaptive form, each
KDE-smoothed into a density so it is scored on likelihood as well as CRPS.

The protocol is a \emph{fair rolling one-step-ahead}: each baseline is fit on an
expanding or rolling window with periodic refit, and every method --- ours and
theirs --- is turned into the \emph{same} \code{Dist} and scored on held-out
log-likelihood \emph{and} CRPS over the same test window. Parametric Student-$t$
predictives (\pkg{arch} and \pkg{rugarch} \pkg{GARCH}, \pkg{ADAM}) are represented
as finite Gaussian scale mixtures matching the analytic $t$ density to $\sim
10^{-3}$; sample- or quantile-based emitters (\pkg{nnetar}, CSP) are
kernel-smoothed into the same \code{Dist}. We report per-series win-rates (the
fraction of series on which \code{laplace} scores better) on the $5{,}402$
continuous series --- those with under $5\%$ exactly-repeating one-step changes --- since on repeating/grid series
the lattice projection (Section~\ref{sec:lattice}) gives a large but
metric-specific advantage that would otherwise dominate the aggregate. Because
many \proglang{FRED} families are panels (a yield curve, an index family) that are
far from independent, we report each win-rate both \emph{raw} and
\emph{family-weighted} (one vote per leading-letter family). Table~\ref{tab:sota}
reports the result (the unified \code{study.py sota} run with
\code{STUDY\_EXCLUDE\_PRICE}).\footnote{The tables in this paper use the default at
submission, \code{scale\_alpha}${=}0.01$ (the terminal leaf's residual-variance
smoothing rate). The shipped default was subsequently raised to $0.03$, which tracks
changing volatility faster; validated out-of-sample on the continuous non-price
universe this improves \code{laplace}'s held-out log-likelihood (${\sim}{+}0.01$ to
${+}0.02$ nats, ${\sim}79\%$ of series) \emph{and} its CRPS (${\sim}80\%$ of series)
uniformly, on both non-price and price series, nudging every win-rate up by roughly a
point or two --- it strengthens the reported figures rather than altering the picture.
Every table reproduces exactly with \code{laplace(scale\_alpha=0.01)}.}

\begin{table}[t!]
\centering
\begin{tabular}{lcccc}
\toprule
Baseline & LL (raw / fam.) & CRPS (raw / fam.) & Mean LL & $N$ \\
\midrule
\pkg{AutoARIMA} (\pkg{statsforecast})  & \textbf{87 / 87\,\%} & 58 / 63\,\% & $1.18$ & $1688$ \\
\pkg{AutoETS} (\pkg{statsforecast})    & \textbf{95 / 94\,\%} & \textbf{77 / 78\,\%} & $1.26$ & $1688$ \\
\pkg{auto.arima} (\proglang{R})        & \textbf{84 / 85\,\%} & 51 / 60\,\% & $1.15$ & $1918$ \\
\pkg{thetaf} (\proglang{R})            & \textbf{92 / 95\,\%} & \textbf{74 / 79\,\%} & $1.24$ & $1918$ \\
\pkg{ADAM} (\proglang{R})              & \textbf{98 / 97\,\%} & \textbf{80 / 82\,\%} & $0.98$ & $1918$ \\
\pkg{nnetar} (\proglang{R})            & \textbf{97 / 95\,\%} & \textbf{64 / 63\,\%} & $-4.56$ & $1918$ \\
\pkg{GARCH}$(1,1)$-$t$ (\pkg{arch})    & $51 / 50\,\%$ & $46 / 48\,\%$ & $1.69$ & $1742$ \\
\pkg{GARCH}$(1,1)$-$t$ (\pkg{rugarch}) & \textbf{90 / 91\,\%} & \textbf{60 / 62\,\%} & $0.27$ & $1918$ \\
\pkg{AutoARIMA} + conformal            & \textbf{82 / 80\,\%} & 36 / 49\,\% & $0.28$ & $1685$ \\
\pkg{AutoARIMA} + ACI                  & \textbf{84 / 81\,\%} & 36 / 49\,\% & $0.45$ & $1685$ \\
CSP                                    & \textbf{98 / 97\,\%} & \textbf{98 / 97\,\%} & $-0.30$ & $5402$ \\
CSP-adaptive                           & \textbf{98 / 97\,\%} & \textbf{98 / 97\,\%} & $-0.09$ & $5402$ \\
\bottomrule
\end{tabular}
\caption{Per-series win-rate of \code{laplace} versus each baseline on the
\textbf{non-price} \proglang{FRED} universe --- $5{,}402$ continuous daily
change-series ($1{,}213$ correlated families), from a $7{,}607$-series sweep
spanning daily, weekly, and monthly frequencies. Slower baselines cover a
per-method subset under their compute budget (column $N$); \code{laplace} and CSP
cover all $5{,}402$. Win-rates are \emph{raw} (per series) and
\emph{family-weighted} (one vote per leading-letter family,
Section~\ref{sec:ablations}). LL counts series where \code{laplace} has the higher
held-out log-likelihood, CRPS where it has the lower CRPS; bold marks a
\code{laplace} win ($>50\%$). ``Mean LL'' is the \emph{baseline's} mean held-out
log-likelihood (\code{laplace}: $\mathbf{1.56}$; higher is better). Only
\pkg{GARCH}-$t$ (\pkg{arch}) escapes a \code{laplace} per-series likelihood win
--- a $50/50$ coin-flip that Section~\ref{sec:martingality} resolves by
martingality.}
\label{tab:sota}
\end{table}

\begin{table}[t!]
\centering
\small
\setlength{\tabcolsep}{5pt}
\begin{tabular}{lrrrc}
\toprule
Baseline & median $\Delta$ & mean $\Delta$ & fam.\ mean & $95\%$ CI \\
\midrule
\pkg{AutoARIMA} (\pkg{statsforecast}) & $+0.17$ & $+0.63$ & $+0.80$ & $[+0.68, +0.93]$ \\
\pkg{AutoETS} (\pkg{statsforecast})   & $+0.21$ & $+0.55$ & $+0.59$ & $[+0.51, +0.67]$ \\
\pkg{auto.arima} (\proglang{R})       & $+0.15$ & $+0.58$ & $+0.76$ & $[+0.65, +0.88]$ \\
\pkg{thetaf} (\proglang{R})           & $+0.18$ & $+0.50$ & $+0.57$ & $[+0.49, +0.65]$ \\
\pkg{ADAM} (\proglang{R})             & $+0.27$ & $+0.75$ & $+0.84$ & $[+0.74, +0.94]$ \\
\pkg{nnetar} (\proglang{R})           & $+1.60$ & $+6.29$ & $+7.88$ & $[+6.97, +8.90]$ \\
\pkg{GARCH}$(1,1)$-$t$ (\pkg{arch})   & $+0.003$ & $+0.11$ & $+0.05$ & $[+0.00, +0.11]$ \\
\pkg{GARCH}$(1,1)$-$t$ (\pkg{rugarch})& $+0.38$ & $+1.47$ & $+1.49$ & $[+1.27, +1.75]$ \\
\pkg{AutoARIMA} + conformal            & $+0.42$ & $+1.53$ & $+1.82$ & $[+1.55, +2.12]$ \\
\pkg{AutoARIMA} + ACI                  & $+0.37$ & $+1.36$ & $+1.72$ & $[+1.44, +2.04]$ \\
CSP                                    & $+0.76$ & $+1.86$ & $+1.87$ & $[+1.66, +2.11]$ \\
CSP-adaptive                           & $+0.75$ & $+1.65$ & $+1.67$ & $[+1.51, +1.86]$ \\
\bottomrule
\end{tabular}
\caption{Paired held-out log-score differences,
$\Delta = \overline{\log p_{\code{laplace}}} - \overline{\log p_{\text{baseline}}}$
per series, on the continuous non-price universe (same coverage as
Table~\ref{tab:sota}). ``fam.\ mean'' weights one vote per leading-letter
family; the interval is a $2{,}000$-draw block bootstrap over families. Paired
differences are scale-invariant (the Jacobian of the change transform cancels
within a series), unlike the absolute ``Mean LL'' column of
Table~\ref{tab:sota}. Every median and every family-weighted mean is positive;
against the \pkg{arch} \pkg{GARCH}-$t$ the median is $+0.003$ --- a genuine
near-tie, resolved along the martingality axis in
Section~\ref{sec:martingality}.}
\label{tab:paired}
\end{table}

\paragraph{The likelihood race.}
On the non-price universe \code{laplace} wins the per-series held-out
log-likelihood race against \emph{eleven of the twelve} baselines by a wide
margin --- the \pkg{ARIMA} family (both \pkg{statsforecast} and the real
\proglang{R} \pkg{auto.arima}), the smoothing and state-space models
(\pkg{AutoETS}, \pkg{thetaf}, \pkg{ADAM}), the neural autoregression
(\pkg{nnetar}), the \pkg{rugarch} \pkg{GARCH}-$t$, and both conformal families ---
typically on $82$--$98\%$ of series, and the win survives family-clustering. The
lone exception is the \pkg{arch} \pkg{GARCH}$(1,1)$-$t$, a $50/50$ tie treated
next. Because win-rates discard magnitude and absolute mean log-likelihoods are
not scale-comparable across mixed-unit series, Table~\ref{tab:paired} adds the
natural aggregate for a proper scoring rule: the \emph{paired} per-series
log-score difference, whose Jacobian term cancels within a series. The paired
view sharpens the claim: the median and family-weighted mean $\Delta$LL are
positive against all twelve baselines, and the \pkg{arch} \pkg{GARCH}-$t$
contest is a genuine near-tie (median $+0.003$, family CI $[+0.00, +0.11]$)
whose apparent mean-LL reversal in Table~\ref{tab:sota} ($1.69$ vs $1.56$,
computed over different coverage) is an artefact of absolute-score averaging
that the paired column resolves --- driven by the near-random-walk tail, which
is exactly the martingality effect of Section~\ref{sec:martingality}.

\paragraph{\pkg{GARCH}-$t$, and the price caveat.}
\pkg{GARCH}$(1,1)$-$t$ is the strongest opponent on log-likelihood, purpose-built
for the heavy tails and volatility clustering of financial change-series. On the
non-price universe it is the only baseline \code{laplace} does not clearly beat on
likelihood --- a $51/50\%$ coin-flip --- and that near-tie hides the real
structure: more than for any other baseline, the \pkg{GARCH}-$t$ contest is governed
almost entirely by how
\emph{martingale-like} the series is, which we make precise in
Section~\ref{sec:martingality}. The other half is the price regime.
Restricting instead to the asset-price classes we
excluded --- equity, FX, commodity returns --- reverses the verdict, and
Table~\ref{tab:price} makes the caveat quantitative rather than narrative. We
make no general-purpose claim on price series and recommend \pkg{GARCH}-$t$
there; this is No-Free-Lunch, reported as a split rather than hidden inside a
pooled average.

\begin{table}[t!]
\centering
\begin{tabular}{lrrrr}
\toprule
Universe & $N$ & LL win & median $\Delta$LL & CRPS win \\
\midrule
non-price continuous            & $1{,}742$ & $51\%$ & $+0.003$ & $46\%$ \\
\quad low-martingality tercile  & $580$     & $78\%$ & $+0.182$ & $78\%$ \\
\quad mid tercile               & $580$     & $44\%$ & $-0.014$ & $42\%$ \\
\quad high-martingality tercile & $582$     & $31\%$ & $-0.030$ & $17\%$ \\
asset price/returns             & $2{,}500$ & $10\%$ & $-0.034$ & $7\%$  \\
\quad with \code{garch\_leaf}   & $2{,}500$ & $18\%$ & $-0.031$ & $5\%$  \\
\bottomrule
\end{tabular}
\caption{\code{laplace} vs the \pkg{arch} \pkg{GARCH}$(1,1)$-$t$, by regime.
``LL win'' is the fraction of series where \code{laplace} has the higher
held-out log-likelihood; $\Delta$LL is the paired per-series log-score
difference. Terciles split the continuous non-price universe by the
martingality score $m$ of Section~\ref{sec:martingality} (low $m$ = mean
structure). The price rows are a $2{,}500$-series stratified sample of the
equity/FX/commodity classes ($34$ correlated families) under the identical
one-step protocol: \pkg{GARCH}-$t$ wins the per-series likelihood race $9$
to $1$, the martingale-heavy extreme of the same axis. Swapping in the
library's \pkg{GARCH}$(1,1)$-$t$ terminal leaf (\code{laplace(leaf=garch\_leaf)})
roughly halves the paired deficit (mean $\Delta$LL $-0.027 \to -0.015$) without
closing it --- and both \code{laplace} variants still beat \pkg{AutoARIMA} and
\pkg{AutoETS} on $68\%$ of the same price series.}
\label{tab:price}
\end{table}

\paragraph{CRPS.}
As ever, CRPS tells a mixed story. \code{laplace} beats the mean-model and
state-space baselines on CRPS --- family-weighted, \pkg{AutoARIMA} $63\%$,
\pkg{AutoETS} $78\%$, \pkg{thetaf} $79\%$, \pkg{ADAM} $82\%$, and the
density-emitting CSP $97\%$ --- but the split tightens against the specialists: it
is roughly even with the \pkg{arch} \pkg{GARCH}-$t$ ($48\%$) and \emph{below} the
fitted-mean conformal variants ($49\%$), which are CRPS-optimised (or
volatility-specialised) by construction. This is
exactly the thesis of Section~\ref{sec:leaf}: likelihood is the metric where a
faithful density wins; CRPS is what the conformal variants are constructed to
optimise and where the volatility specialists are strongest, and there we are
competitive but not dominant. A practitioner who truly only cares about CRPS
can repoint the leaf (Section~\ref{sec:ablations}); a practitioner who cares about
the density --- for risk, sizing, or combination --- should prefer the
log-likelihood column.

\paragraph{Coverage.}
Proper scores reward calibration only implicitly; we also measured it directly.
On $1{,}000$ continuous non-price series (same rolling protocol), the empirical
coverage of each method's \emph{nominal central $90\%$} interval over the
$300$ held-out points: \code{laplace} covers a median $89.3\%$ (mean $88.4\%$)
and is within $\pm2$ points of nominal on $65\%$ of series --- its aggregate
PIT histogram is nearly flat (bin shares $0.046$--$0.059$ against an ideal
$0.050$). The baselines are systematically overdispersed: \pkg{AutoARIMA},
\pkg{AutoETS}, and \pkg{GARCH}-$t$ deliver $93.7$--$95.0\%$ median coverage
from their $90\%$ bands and land within $\pm2$ points on only $12$--$17\%$ of
series. An interval that quietly covers $95\%$ is not conservative comfort; it
is a miscalibrated risk limit. Calibration is also what makes the
probability-integral-transform diagnostics that \pkg{skaters} exposes in its
state (\code{state["pit"]}, \code{state["z"]}) usable as an anomaly monitor:
the thresholds mean what they say
(harness: \code{benchmarks/coverage\_study.py}).

\paragraph{Conformal prediction, specifically.}
The conformal opponents deserve a separate word; two kinds appear here. Conformal
seasonal pools (CSP) sample a seasonal residual pool and KDE-smooth it into a
density, so --- unlike a raw conformal predictive system, whose cumulative
distribution assigns zero density, and hence $-\infty$ log-likelihood, outside the
observed residual range --- CSP \emph{can} be scored on likelihood. \code{laplace}
beats it on both metrics ($97\%$ family-weighted on each), because a recency-blind
seasonal pool tracks neither drift nor the volatility clock. Against the stronger
\emph{fitted-mean} conformal variants (\pkg{AutoARIMA}-mean residuals,
split-conformal and ACI) the split is clean: \code{laplace} has the
higher held-out log-likelihood on the large majority ($80$--$81\%$
family-weighted), while the conformal variants roughly match it on CRPS
($\sim 49\%$). That is exactly as it should be --- finite-sample marginal coverage is
conformal's actual guarantee, CRPS the score its interval-like output does well
on, and it has no density to surrender on likelihood --- and it is the
empirical face of the pluggable-base argument: the
right objective for \emph{modelling} is the density, and the one place a
downstream score should get a vote is the terminal leaf where the conjugation
bottoms out. A user who wants CRPS-trained
residual shaping can have it inside the same library by repointing that leaf to
CRPS (Section~\ref{sec:ablations}) --- a CRPS-fit density, not conformal's
finite-sample coverage guarantee --- without giving up the density the conformal
systems cannot provide.

\paragraph{Scope.}
The non-price economic universe ($5{,}402$ continuous change-series of a
$7{,}607$-series sweep spanning daily, weekly, and monthly frequencies), one-step
horizon. The price/return regime is reported as the explicit caveat
above and is not part of the general-purpose claim; multi-step horizons are
treated in Section~\ref{sec:multistep}; raw levels are left for later; zero-shot
foundation models get their own protocol in Section~\ref{sec:foundation}.

\subsection{The martingality gradient} \label{sec:martingality}

The one baseline \code{laplace} does not simply dominate is \pkg{GARCH}-$t$, and
the reason is not the price/non-price label per se but a property that varies
\emph{within} the non-price universe: how close each change-series is to a
martingale. A pure martingale difference has no exploitable conditional mean ---
only its conditional \emph{variance} can be modelled, which is exactly what
\pkg{GARCH} specialises in; a series with mean structure (mean reversion,
momentum) hands a forecaster signal that \pkg{GARCH}'s constant mean discards.

We quantify this with the Lo--MacKinlay variance ratio \citep{lo1988stock},
$\mathrm{VR}(q) = \operatorname{Var}\!\big(\sum_{i=1}^{q}\Delta_{t+i}\big) /
\big(q\operatorname{Var}(\Delta_t)\big)$: $\mathrm{VR}=1$ is a random walk,
$\mathrm{VR}<1$ mean reversion, $\mathrm{VR}>1$ momentum. We score martingality
$m = \exp(-\lvert\ln \mathrm{VR}(5)\rvert) \in (0,1]$, with $m\to 1$ the pure
random walk. Binning the daily-continuous non-price series into martingality
terciles exposes the same axis in \emph{every} contest, with a sign that reveals
what each baseline models (Table~\ref{tab:mart}). The \pkg{GARCH}-$t$ (\pkg{arch})
contest is the extreme: \code{laplace} wins the held-out log-likelihood on $98\%$
of the most mean-predictable series and only $5\%$ of the near-random-walks, so its
aggregate ($\approx 50\%$) is not a close-run tie but the average of a decisive win
and a decisive loss. The gradient is not unique to it, though. It runs the
\emph{same} way, more mildly, for every conditional-variance or trend model --- the
\pkg{rugarch} \pkg{GARCH}-$t$ ($100 \!\to\! 73\%$), \pkg{thetaf} ($98 \!\to\! 69\%$),
\pkg{AutoETS} ($98 \!\to\! 85\%$) --- because a random walk leaves only the variance
to model, their comparative strength. It runs the \emph{other} way for the mean
models: against \pkg{AutoARIMA}, \code{laplace} wins \emph{least} on the
mean-predictable end ($67\%$, where an \pkg{ARIMA} mean genuinely fits) and
\emph{more} toward the random walk ($87\%$). And it is flat, near $100\%$
throughout, for the strong general models (\pkg{ADAM}, \pkg{nnetar}, CSP).
Martingality is thus a general organising axis --- it sorts the opponents by what
they model, variance specialists one way, mean specialists the other, generalists
not at all --- and \pkg{GARCH}-$t$ is simply its sharpest case. The axis is not
an artefact of the variance-ratio statistic. Binning instead by the absolute
lag-one autocorrelation of the changes reproduces the same structure: against
\pkg{GARCH}-$t$, \code{laplace} wins $27\%$ of the least-autocorrelated
tercile and $81\%$ of the most; against \pkg{AutoARIMA} the gradient runs the
other way ($92\%$ down to $78\%$); the generalists stay flat. We note the
variance ratio is a second-order diagnostic --- ``martingale-like'' here means
low second-order mean predictability, not a complete characterisation of the
conditional mean. This is also the
mechanism behind the price caveat: price/return series are the martingale-heavy
extreme of the axis, so the recommendation is precise --- reach for \pkg{GARCH}-$t$
not ``on prices'' but ``on whatever is close to a random walk.''

\begin{table}[t!]
\centering
\begin{tabular}{lccc}
\toprule
Opponent (what it models) & low $m$ & mid & high $m$ \\
\midrule
\pkg{GARCH}-$t$ (\pkg{arch}) --- variance    & \textbf{98} & $42$ & $5$ \\
\pkg{GARCH}-$t$ (\pkg{rugarch}) --- variance & \textbf{100} & \textbf{86} & \textbf{73} \\
\pkg{thetaf} --- trend                        & \textbf{98} & \textbf{92} & \textbf{69} \\
\pkg{AutoETS} --- smoothing                   & \textbf{98} & \textbf{96} & \textbf{85} \\
\midrule
\pkg{ADAM} --- state-space                    & \textbf{100} & \textbf{100} & \textbf{99} \\
\pkg{nnetar} --- neural AR                    & \textbf{100} & \textbf{99} & \textbf{97} \\
CSP, CSP-adaptive --- conformal pool          & \textbf{100} & \textbf{100} & \textbf{100} \\
\pkg{AutoARIMA} + conformal                   & \textbf{82} & \textbf{80} & \textbf{83} \\
\pkg{AutoARIMA} + ACI                         & \textbf{83} & \textbf{86} & \textbf{86} \\
\midrule
\pkg{auto.arima} (\proglang{R}) --- mean      & \textbf{68} & \textbf{85} & \textbf{77} \\
\pkg{AutoARIMA} (\pkg{statsforecast}) --- mean & \textbf{67} & \textbf{93} & \textbf{87} \\
\bottomrule
\end{tabular}
\caption{Per-series held-out log-likelihood win-rate of \code{laplace} (\%) by
martingality tercile, daily-continuous non-price series ($n = 1{,}299$, $\approx
433$ per tercile); $m=\exp(-\lvert\ln\mathrm{VR}(5)\rvert)$ from the Lo--MacKinlay
variance ratio, $m\to1$ = random walk. Martingality modulates \emph{every}
contest, with a sign set by what the baseline models: for conditional-variance and
trend models (top block) \code{laplace}'s edge \emph{shrinks} toward the
random-walk end; for \pkg{ARIMA} mean models (bottom) it \emph{grows}; for the
strong general and conformal opponents (middle) it is flat near $100\%$.
\pkg{GARCH}-$t$ (\pkg{arch}) is the sharpest case. Bold marks a \code{laplace} win
($>50\%$).}
\label{tab:mart}
\end{table}

\paragraph{Robustness of the win.}
The likelihood win is not an artefact of one slice of the universe. It holds
across observation frequency --- \code{laplace} wins the log-likelihood race
against every baseline on daily, weekly, \emph{and} monthly non-price series --- and
across sequence length (from the $200$-change minimum upward) and stickiness (the
fraction of exactly-repeating, zero-change steps, where the lattice projection of
Section~\ref{sec:lattice} contributes). An interactive companion recomputes every
per-series win-rate and the accuracy/speed frontier live over any user-selected
subset --- sampled at random, balanced by category, or filtered by frequency,
length, stickiness, or martingality --- at
\url{https://skaters.microprediction.org/robustness.html}.

\subsection{Multi-step horizons and the granularity mixture} \label{sec:multistep}

%% MULTISTEP-NUMBERS: full non-price universe (benchmarks/multistep_study.py);
%% single-scale rows = method "laplace" (scales=[1]); multi-scale rows =
%% method "laplace-msw" (the shipped default, evaluated online).
Multi-step is where the granularity axis of Section~\ref{sec:granularity}
earns its place. The target is stated exactly: at origin $t$ and horizon $h \in
\{1, 5, 20\}$, every forecaster emits a predictive for $\Delta_{t+h}$ --- the
\emph{single} one-period change realised $h$ steps later --- not the cumulative
$h$-period change and not the level. Each horizon's predictive is scored as the
same \code{Dist} on held-out log-likelihood and CRPS at $24$ evenly spaced
origins per series. The opponents are the baselines for which a multi-step
predictive is principled rather than improvised: \pkg{AutoARIMA} and
\pkg{AutoETS} (\code{forecast(h)} from a periodically refit window),
\pkg{GARCH}$(1,1)$-$t$ (its analytic $h$-step variance ladder), and
\pkg{Prophet} \citep{taylor2018prophet}, whose fit-once-predict-$h$ design this
protocol matches exactly. The conformal variants are one-step constructions and
are excluded.

\begin{table}[t!]
\centering
\begin{tabular}{llcccc}
\toprule
Method & $h$ & \pkg{AutoARIMA} & \pkg{AutoETS} & \pkg{GARCH}-$t$ & \pkg{Prophet} \\
\midrule
single-scale       & $1$  & \textbf{75\,\%} / \textbf{58\,\%} & \textbf{81\,\%} / \textbf{69\,\%} & \textbf{66\,\%} / \textbf{59\,\%} & \textbf{87\,\%} / \textbf{75\,\%} \\
                   & $5$  & \textbf{72\,\%} / \textbf{55\,\%} & \textbf{77\,\%} / \textbf{68\,\%} & \textbf{59\,\%} / 45\,\% & \textbf{76\,\%} / \textbf{58\,\%} \\
                   & $20$ & \textbf{60\,\%} / 46\,\%          & \textbf{66\,\%} / \textbf{59\,\%} & 50\,\% / 36\,\%          & \textbf{63\,\%} / 43\,\% \\
\midrule
multi-scale        & $5$  & \textbf{80\,\%} / \textbf{62\,\%} & \textbf{85\,\%} / \textbf{74\,\%} & \textbf{62\,\%} / 49\,\% & \textbf{81\,\%} / \textbf{65\,\%} \\
(the default)      & $20$ & \textbf{73\,\%} / \textbf{53\,\%} & \textbf{79\,\%} / \textbf{64\,\%} & \textbf{60\,\%} / 38\,\% & \textbf{72\,\%} / 50\,\% \\
\bottomrule
\end{tabular}
\caption{Multi-step per-series win-rates (LL / CRPS) of \code{laplace} on
$7{,}335$ series of the non-price universe, $24$ origins per series. Rows
contrast the single-scale variant (\code{scales=[1]}, the native fan-out) with
the multi-scale mixture that is the default at $k>1$
(Section~\ref{sec:granularity}); at $h=1$ the two coincide by construction and
the multi-scale row is omitted. Bold marks a win ($>50\%$). The single-scale
fan-out decays with horizon --- mean held-out log-likelihood $+2.65$, $+2.44$,
$+1.10$ at $h = 1, 5, 20$ --- while the granularity mixture holds it ($+2.70$,
$+2.53$ at $h = 5, 20$), lifts every $h\ge 5$ log-likelihood contest by
$3$--$13$ points, and beats the single-scale variant head-to-head on $68\%$ of
series at $h=20$ ($59\%$ at $h=5$; exact ties at $h=1$).}
\label{tab:multistep}
\end{table}

Three findings (Table~\ref{tab:multistep}). First, the native fan-out decays
with horizon: \code{laplace} wins the likelihood race against everything at
$h=1$ but slides toward a coin-flip against \pkg{GARCH}-$t$ by $h=20$, and its
mean log-likelihood drops from $+2.65$ to $+1.10$ as the $k$-step variance
inflates --- severely so on a small set of series that dominate the (therefore
unreported) mean CRPS. Second, a \emph{fixed} coarse clock is not the answer:
decimating to stride $h$ and forecasting one step rescues the $h=20$ density
(mean $+2.43$) but throws away the intermediate points and loses to the mixture
on $72\%$ of series at $h=20$ ($75\%$ at $h=5$). Third, the likelihood-weighted
mixture of scales recovers both ends: it beats the native fan-out on $68\%$ of
series at $h=20$, leaves $h=1$ untouched by construction (the wrapper passes
scale one through), and wins every $h\ge5$ log-likelihood contest. The gain is flat across autocorrelation terciles ($63/59/63\%$ vs
native), consistent with its mechanism: it repairs the variance fan-out rather
than exploiting mean structure. Granularity behaves like the other candidate
axes --- mixed under likelihood weights, the wrong scale costs only its softmax
share.

\subsection{Against zero-shot foundation models} \label{sec:foundation}

Pretrained time-series foundation models use a fundamentally different protocol:
they are applied \emph{zero-shot}, conditioning on a context window with no
fitting. We evaluate four ---
Chronos-Bolt \citep{ansari2024chronos}, TimesFM~2.5 \citep{das2024timesfm},
Moirai \citep{woo2024moirai}, and Lag-Llama \citep{rasul2023lagllama} --- on 120
change-series (69 continuous). At each step the model receives a fixed
$256$-length window of the preceding changes and predicts the next change, with
all windows for a series batched into one inference call. Each predictive is
turned into the same \code{Dist} --- the native Student-$t$ head for
Moirai/Lag-Llama, a sample/quantile reconstruction for Chronos-Bolt/TimesFM ---
and \code{laplace} is re-scored on the identical window. Table~\ref{tab:foundation}
reports per-series win-rates.

\begin{table}[t!]
\centering
\begin{tabular}{llccc}
\toprule
Model & density & LL (all / cont.) & CRPS (all / cont.) & Mean LL (cont.) \\
\midrule
Moirai (1.1-R-small)  & native $t$    & 98 / \textbf{97}\,\% & 94 / 93\,\% & 0.92 \\
Lag-Llama             & native $t$    & 67 / \textbf{99}\,\% & 65 / 94\,\% & 0.39 \\
Chronos-Bolt (small)  & quantile$^{*}$ & 76 / \textbf{100}\,\% & 67 / 88\,\% & $-1.96$ \\
TimesFM (2.5-200M)    & quantile$^{*}$ & 75 / \textbf{100}\,\% & 58 / 72\,\% & $-1.18$ \\
\bottomrule
\end{tabular}
\caption{Per-series win-rate of \code{laplace} versus four zero-shot foundation
models on 120 change-series (69 continuous). $^{*}$Chronos-Bolt/TimesFM
log-likelihood is reconstructed from a quantile head and is tail-limited; read
their CRPS as the fairer signal. \code{laplace} mean continuous logpdf:
1.51.}
\label{tab:foundation}
\end{table}

On the continuous series \code{laplace} beats all four foundation models
zero-shot on both metrics (97--100\% LL, 72--94\% CRPS). The native-density
models (Moirai, Lag-Llama) are the meaningful likelihood opponents and the
closest, but still lose per-series. On repeat-heavy/grid series the foundation
models are competitive or better --- Lag-Llama even beats \code{laplace} on mean
log-likelihood over the \emph{full} universe ($6.54$ vs $4.52$) by placing tight
mass on revisited values, the same effect as the lattice projection --- which is
why the continuous split matters. We emphasise in the main text, not as a
footnote, that this is a \emph{narrow} comparison and not a broad
foundation-model verdict: zero-shot, no refit, a fixed $256$ context, and
forecasting differenced streams rather than the levels these models were
chiefly trained on --- a protocol chosen because it is the one a practitioner
with a single stream and no fitting budget actually faces.\footnote{One
exploratory check: naively fine-tuning Lag-Llama on a single series' history
(5 epochs) overfit badly, collapsing held-out log-likelihood from $+1.5$ to
below $-100$ at $\sim$15$\times$ the cost. We draw no general conclusion about
fine-tuning from a single naive configuration --- adapting a pretrained model to
one short univariate stream is outside its design regime, and domain-level
fine-tuning across many series remains untested here; the zero-shot result is
the headline because zero-shot is the intended use.}

%% -------------------------------------------------------------------------
\section{Ablations: repointing the objective} \label{sec:ablations}

A companion sweep on $10{,}822$ \proglang{FRED} series (every daily-tagged series
by popularity) isolates the effect of the terminal-leaf objective against the
\pkg{crepes} conformal predictive system (scored on its own CDF via the pinball
decomposition of CRPS). \code{laplace} runs in its fixed default configuration on
every series --- no per-series tuning --- while the opponent gets its best
calibration window per series. Because this conformal opponent uses a naive
(zero-change) mean, we read it as an ablation of our own objective knob rather
than a state-of-the-art claim. Per-policy, de-correlating to one vote per
correlated family (198 families):

\begin{table}[h!]
\centering
\begin{tabular}{lccc}
\toprule
Forecaster (identical structure) & CRPS, raw & CRPS, family & Mean LL \\
\midrule
likelihood trunk + likelihood leaf & 92.1\,\% & 55.6\,\% & 2.97 \\
likelihood trunk + CRPS leaf       & \textbf{96.6\,\%} & \textbf{64.8\,\%} & \textbf{2.99} \\
bare CRPS leaf (no trunk)          & 92.6\,\% & 61.0\,\% & 2.82 \\
\bottomrule
\end{tabular}
\caption{Effect of the terminal-leaf objective (fixed \code{laplace} vs
best-of-crepes, $10{,}822$ series). The middle row is the \code{laplace} default
(CRPS leaf, lattice projection on). A general likelihood policy already edges conformal on CRPS over independent
families ($55.6\%$); repointing only the terminal leaf to CRPS lifts that to
$64.8\%$ while slightly raising mean held-out log-likelihood in this sweep
($2.97 \rightarrow 2.99$; a small difference we read as ``no likelihood cost'',
not as a win), matching the dedicated CRPS specialist while modelling
faithfully. Crepes' CRPS is U-shaped in
window length (optimum near 250); \code{laplace} beats its best-of-$\{60\dots1500\}$
on $96.4\%$ of series, so the gap is structural, not a calibration-window choice.}
\label{tab:objective}
\end{table}

Three further observations motivate the small public API. Priors wash out:
across $25$ series with $\geq 2000$ changes the conventional prior-flavoured
policies (trend, long-memory, drift, minimax, fast/slow) sit within $0.0006$ nats
of the uniform-prior general forecaster on the second half of each series, and the
first-half winner beats the general forecaster on the second half only $44\%$ of
the time, worse than chance. Conform-last is near-free: on idealised light-tailed
processes the CRPS leaf costs $0.005$--$0.007$ nats versus the likelihood leaf,
while on heavy-tailed and stochastic-volatility processes it helps (up to
$+0.04$); real data lives in the second regime. And the lattice projection is free
when unused. Together these justify exposing exactly one general forecaster
(uniform prior, CRPS leaf, lattice projection and coordinate learning on by
default), with volatility and mean-reversion left to composable transforms rather
than separate named models; the remaining structural ideas do not pay their way
out of sample.

%% -------------------------------------------------------------------------
\section{Discussion} \label{sec:discussion}

The unifying theme is No-Free-Lunch \citep{wolpert1997no} taken seriously. Every
``method'' is a prior; averaged over all series none dominates. The correct
response is not to choose one but to make the priors explicit, weight them online,
and bound the cost of a wrong prior by how fast the weighting abandons it. Read
that way, the named methods of classical forecasting are convenience bundles, and
the natural object is a single adaptive ensemble with two orthogonal terminal knobs
--- \emph{which proper scoring rule the leaf optimises}, and \emph{whether a
lattice projection is active} --- neither of which is a prior over the trunk. The
granularity mixture of Section~\ref{sec:granularity} is the same argument applied
to the sampling clock: which time scale a forecaster commits to is one more
implicit prior, and the multi-step results show it too is better weighted online
than chosen.

\paragraph{Anomaly detection as future work.}
A calibrated online density makes anomaly detection nearly a by-product. The
state exposes, at every horizon $m \le k$, the probability integral transform
of each arriving point under the predictive issued $m$ steps earlier
(\code{state["pit"]}, approximately Uniform when the stream behaves) and its
standard-normal image (\code{state["z"]}), so a threshold on $|z|$ is an alarm
whose false-positive rate means what it says --- the coverage results above are
precisely the license --- and the multi-horizon structure separates one-off
spikes (surprising at $m=1$, forgotten by $m=20$) from genuine breaks
(surprising at every matured horizon at once). Because the model has already
absorbed drift, the volatility clock, seasonality, and coordinate, what remains
surprising is surprising relative to an adapted forecast rather than to a
rolling window. We deliberately stop short of a benchmark claim, for two reasons.
First, the definition itself is unsettled: an anomaly is only ever ``a pattern
that does not conform to expected behaviour'' \citep{chandola2009anomaly},
which makes the concept relative to a model of expectation and a purpose ---
our $|z|$ is precise about \emph{whose} expectation (the adapted predictive's),
but whether that coincides with what a benchmark's human labeller meant is
exactly the unresolved part. Second, the empirical evaluation is notoriously
fraught --- \citet{wu2023current} argue that the prevailing benchmark suites
are flawed to the point of ``creating the illusion of progress,'' and
\citet{kim2022rigorous} show that a common scoring adjustment lets even random
scores appear state-of-the-art --- and our own preliminary experiments on the Numenta
benchmark \citep{lavin2015nab} indicate that leaderboard position there is
dominated by detector-layer machinery (per-stream score normalisation and
thresholding economics) rather than by the quality of the underlying
predictive. Turning a calibrated surprise stream into a benchmark-competitive
detector, and finding an evaluation worth competing on, is future research. The
empirical headline supports the design: against the strongest distributional
opponents available on CPU, including the heavy-tail and neural state of the art,
the general forecaster wins the likelihood race, and where it does not lead ---
CRPS --- the deficit is confined to methods explicitly optimised for that metric
and is recoverable by repointing a single knob.

%% -------------------------------------------------------------------------
\section{Heritage and reproducibility} \label{sec:heritage}

\pkg{skaters} is a from-scratch rewrite distilling ideas from \pkg{timemachines}
\citep{cotton2021timemachines}, an earlier competition-tested online-forecasting
package, and builds on years of live distributional-prediction contests run on
the microprediction platform \citep{cotton2022microprediction}, where forecasts
were scored continuously as full distributions. The conjugation recursion has a
direct antecedent there: the platform's \emph{multi-level distributional
prediction} routed each stream
into derived residual streams --- the so-called \emph{z-streams} --- that were
themselves predicted distributionally and composed back, ``transforms all the way
down'' avant la lettre; \pkg{skaters} distils that platform mechanism into a
single online, dependency-free abstraction. The library is implemented twice --- pure \proglang{Python} and
zero-dependency \proglang{JavaScript} --- and a parity suite checks roughly
$100{,}000$ probe values to $10^{-6}$ on every run, so results are reproducible
across both runtimes and in the browser via \pkg{Pyodide} \citep{pyodide2021}.
The benchmark of Section~\ref{sec:experiments} is a single parallel harness
(\code{benchmarks/study.py}, invoked as \code{study.py sota}) that scores every
method through the identical \code{Dist} interface; given a \proglang{FRED} key
and the optional baselines (\pkg{statsforecast}, \pkg{arch}, the \proglang{R}
\pkg{forecast}/\pkg{smooth}/\pkg{rugarch} stack, and pure-\proglang{numpy} CSP) it
reproduces Table~\ref{tab:sota} end to end, and
\code{study.py sota summarize} regenerates the reported win-rates from the
committed result file. The multi-step results of Section~\ref{sec:multistep} have
their own harness (\code{benchmarks/multistep\_study.py}) under the same
\code{Dist} protocol. All results in this paper were produced with the
v0.11 defaults; from v0.12 the terminal leaf's scale-tracking rate default
changed (\code{scale\_alpha}: $0.01 \to 0.03$, a small out-of-sample
improvement on both metrics), so pass \code{scale\_alpha=0.01} to reproduce
the reported numbers exactly.

%% -------------------------------------------------------------------------
\bibliographystyle{plainnat}
\bibliography{skaters}

\end{document}
